---
title: "Data Analysis on Salary of Data Engineer"
format: 
  html:
    theme: united
    toc: true
editor: visual
---

```{r}
#| message: False
#| warning: False
#| echo: False
library(targets)
library(dplyr)
library(knitr)
library(kableExtra)
```

# Introduction

We live in a world where every piece of data significantly influences human behavior. Consequently, data scientists and data analysts are essential for interpreting and generating insightful information, enabling organizations to make data-driven decisions. However, data originates from diverse sources with varying formats and structures, complicating immediate utilization. This challenge highlights the vital role of data engineers. They design and implement systems that collect, manage, and transform raw data into formats suitable for analysis. Positioned at the initial step of the data analysis pipeline, the role of a data engineer is indispensable. This is why data engineers are in high demand across various sectors like: finance, retail and technology.

In this report, we will examine the salary trends for data engineers and analyze other factors influencing their income over the past four years from 2020 to 2023. 

The data used for this analysis is [2023 Data Scientists Salary](https://www.kaggle.com/datasets/henryshan/2023-data-scientists-salary) from [Kaggle](https://www.kaggle.com/), whose author is Henry Shan.  

# Initial Data Analysis

Before commencing any analysis, we will review the metadata. The table below highlights every available variables in the data, as well as giving a short interpretation of each variable.

| Variable | Explanation |
|---|---|
| work_year | The year the salary was paid. |
| experience_level | The experience level in the job during the year. <br> **EN**: Entry-level / Junior <br> **MI**: Mid-level / Intermediate <br> **SE**: Senior-level / Expert <br> **EX**: Executive-level / Director |
| employment_type | The type of employment for the role. <br> **PT**: Part-time <br> **FT**: Full-time <br> **CT**: Contract <br> **FL**: Freelance |
| job_title | The role worked in during the year. |
| salary | The total gross salary amount paid. |
| salary_currency | The currency of the salary paid as an ISO 4217 currency code. |
| salaryinusd | The salary in USD. |
| employee_residence | Employee's primary country of residence during the work year as an ISO 3166 country code. |
| remote_ratio | The overall amount of work done remotely. |
| company_location | The country of the employer's main office or contracting branch. |
| company_size | The median number of people that worked for the company during the year. |

## Data format check

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: tbl-dataformat
#| tbl-cap: "Data format"
tar_read(data_class)
```

At the first glance from @tbl-dataformat, variables such as: `experience_level`, `employment_type`. `job_title`, etc. are initially formatted as *character*, which makes the summary statistics generated for these variables meaningless.

To deal with this, we will convert all *character* variables into *factor* variables, then re-generate the data summary statistics.

## Missing values

We then examine the data for any missing values.

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: tbl-missingvalues
#| tbl-cap: "Missing values"
tar_read(missing_data_fac)
```

As shown in @tbl-missingvalues, the data is complete, with no missing values present.

## Filtering data engineer role
```{r}
#| message: False
#| warning: False
#| echo: False

tar_read(unique_job_title)
```

There are several specific data engineer role in the data. However, for this analysis, we will only concentrate on data engineer role in general. And therefore, we will filter out *Data Engineer* value from `job_title`.

## Re-group company_location variable

Moreover, we will re-group `company_location` values to **US** and **Other** for later comparison of salary between US organisations and foreign organisations.

## Filtering full time role

Finally, we then filter from `employment_type` to *Full time* role only.

# Exploratory Data Analysis (EDA)

## Summary statistics

We use `summary()` to get the summary statistics of all variables. Moreover, since we have filtered to only one value for `job_title` and `employment_type`, we will not perform summary statistics for these 2 variables.

### Numerical variables

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: tbl-numsum
#| tbl-cap: "Summary statistics of numerical variables"
tar_read(sum_numeric)
```

Given that the salary for the data engineer role is paid in the currency of the company's location, we will use `salary_in_usd` for our analysis.

As shown in @tbl-numsum, the annual salaries for data engineers range from *12,000 USD* to *324,000 USD*, with a mean salary of approximately **143,000 USD**. This significant disparity in pay can be attributed to various factors, including the location of the role and the data engineer's level of experience.

### Factor variables

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: tbl-facsum
#| tbl-cap: "Summary statistics of factor variables"
tar_read(sum_factor)
```

As we can see from @tbl-facsum, for both `company_loc` as well as `employee_residence`, highest frequency values are *US*. This may happen due to a few reasons:

1. **Large Tech Industry Presence**:
   The United States is home to many of the world's largest technology companies, such as Google, Apple, Microsoft, and Amazon. These companies often have significant demand for data engineers and other data-related roles, leading to a higher concentration of such jobs in the US.

2. **Higher Salaries and Better Opportunities**:
   Salaries for tech roles, including data engineers, tend to be higher in the US compared to other countries. This attracts talent from around the world to seek employment in the US, contributing to the higher frequency of employees residing in the US.

3. **Headquarters Location**:
   Many multinational tech companies have their headquarters in the US. As a result, their main offices are located there, increasing the frequency of the US as the company location.
   
Moreover, most of the data engineers for this data is currently in *SE* or Senior-level role.

## Average Salary Trend

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: fig-salarytrend
#| fig-cap: "Average salary trend"
tar_read(salary_trend_plot)
```


As we can see from @fig-salarytrend, there is a consistent upward trend in the average salary for data engineers from 2020 to 2023.

The average salary has increased significantly each year, indicating a growing demand for data engineers and possibly an increasing recognition of the value they bring to organizations.

## Average Salary over Years by Company Location

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: fig-salarycom
#| fig-cap: "Average salary by company location"

tar_read(salary_company_loc_plot)
```

From @fig-salarycom, we can see that there is a huge gap in salary of data engineers between US organisations and foreign organisations. This may reflect the higher cost of living in the US, greater demand for tech talent, or more competitive salary packages offered by US companies.

* **US**: Salaries in the US have shown a more stable and slightly upward trend overall. However, there is a slight decrease from 2020 to 2021, followed by a significant increase from 2021 to 2022, and a continued rise into 2023.

* **Other Locations**: Salaries in other locations have followed a similar pattern. There is a slight increase from 2020 to 2021, a significant jump from 2021 to 2022, and a continued rise from 2022 to 2023.


## Average Salary over Years by Experience Level

```{r}
#| message: False
#| warning: False
#| echo: False
#| label: fig-salaryexper
#| fig-cap: "Average salary by experience"

tar_read(salary_experience_plot)
```

- **Experience Level Impact**:
  - **Executive-level (EX)**: Executive-level data engineers see the highest salaries, with a notable peak around 2022, indicating high rewards for top-tier experience and leadership roles.
  - **Senior-level (SE)**: Senior-level salaries show a steady upward trend, reflecting the growing value of experienced professionals.
  - **Mid-level (MI)**: Mid-level salaries have been relatively stable with a slight increase, indicating moderate growth in compensation for this experience tier.
  - **Entry-level (EN)**: Entry-level salaries have seen a substantial increase from 2020 to 2022, though there is a slight decline in 2023. This could suggest a correction or stabilization in entry-level hiring and compensation.

## Overall Insights

### Growth in Salaries
Salaries for data engineers have been on the rise overall, with significant variations depending on both location and experience level.

### Location Impact
The United States stands out as the highest-paying region for data engineers. This can be attributed to the country's strong tech industry and higher cost of living, which drives up salary levels.

### Experience Matters
Experience plays a crucial role in determining salary. Executive and senior-level positions earn the highest pay, reflecting their expertise and leadership responsibilities. Entry-level positions, while showing positive growth, exhibit some fluctuations.

These trends underscore the dynamic and evolving landscape of the data engineering field. There is growing recognition of the value data professionals bring, coupled with a competitive job market that continues to push salaries upward.

# Regression Model 

For better understanding on how experience level and company location affects the salary of data engineer, we fit a linear model for `salary_in_usd` against `company_loc` and `experience_level`. We also include `work_year` in the regression as explanatory variable to encounter for the trend pattern.

```{r}
tar_read(summary_lm)
```


### Coefficients and Their Interpretation
* **Intercept**: The negative intercept (-10,086,485) is not directly interpretable in a meaningful way.
  
* **Work Year**: The positive coefficient for `work_year` (5012) indicates that for each additional year, the salary increases by approximately $5,012, holding other factors constant. This suggests a general upward trend in salaries over time.

* **Experience Level**:
  * **Executive Level (EX)**: With a coefficient of 102,667, executive-level data engineers earn significantly higher salaries compared to the baseline (entry-level), all else being equal.
  * **Senior Level (SE)**: The coefficient of 48,895 shows that senior-level engineers earn about $48,895 more than entry-level engineers.
  * **Mid-Level (MI)**: The coefficient of 22,005 indicates that mid-level data engineers earn about $22,005 more than entry-level engineers.


* **Company Location**
  * **US Location**: The coefficient of 57,474 for `company_locUS` means that data engineers working in the US earn about $57,474 more than their counterparts in other locations.

### Statistical Significance
* The p-values for `work_year`, `experience_level` (EX, MI, SE), and `company_locUS` are all less than 0.05, indicating that these predictors are statistically significant.

### Model Fit
* **R-squared**: The model explains about 31.7% of the variance in salaries (`R-squared = 0.317`), which is reasonable for a model with a few predictors but suggests there are other factors influencing salary that are not captured by this model.

### Key Takeaways
1. **Yearly Salary Increase**: There is a steady increase in data engineer salaries over the years, with an average annual increase of $5,012.
2. **Experience Level Impact**: Higher experience levels (especially executive and senior levels) significantly boost salaries, reflecting the premium placed on experience and expertise in the field.
3. **Geographical Impact**: Working in the US significantly enhances salary prospects, likely due to the strong tech industry and higher cost of living.

# Conclusion

### Conclusion

In conclusion, the analysis reveals a significant upward trend in data engineer salaries over the past few years, driven by factors such as experience level and geographical location. Executive and senior-level positions command the highest salaries, highlighting the value placed on experience and expertise. Additionally, data engineers based in the United States enjoy higher salaries compared to those in other locations, reflecting the strong demand and competitive compensation in the US tech industry.

These findings underscore the dynamic nature of the data engineering field and the growing recognition of the critical role data professionals play in modern organizations. As the demand for skilled data engineers continues to rise, understanding the factors that influence salary trends can provide valuable insights for both employers and job seekers, guiding strategic decisions in recruitment, career planning, and compensation management.